Bayesian model comparison for rare variant association studies of multiple phenotypes

### 2018-01-30 tex/manuscript.tex
final changes

 To define the alternative model, we must specify the prior correlation structure, scale, and location (direction) of the effect sizes. Let $N$ be the number of individuals and $K$ the number of phenotype measurements on each individual. Let $M$ be the number of variants in a testing unit ${\bs G}$, where ${\bs G}$ can be, for example, a gene, pathway, or a network. Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size. When considering multiple phenotypes ($K > 1$), we define the prior correlation structure of the effect sizes as an $SMK\times SMK$ matrix $\mathbf{U}$. In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices
\begin{itemize}
\item an $S\times S$ matrix $\mathbf{R_{\textrm{study}}}$ containing the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation}
\item an $M\times M$ matrix $\mathbf{R_{\textrm{var}}}$ containing the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}
\item a $K\times K$ matrix $\mathbf{R_{\textrm{phen}}}$ containing the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}
\end{itemize}
The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimates (or log odds ratios), and the standard errors of the effect size estimates (Online Methods).

@@

Similarly, we can utilize a prior on the location (direction) of effects to specify alternative models where we seek to identify variants with protective effects against disease. Thus far we have assumed that the prior mean, or location, of genetic effects is zero which makes it feasible to analyze a large number of phenotypes without enumerating the prior mean across all phenotypes. To proactively identify genetic variants that have effects that are consistent with a protective profile for a disease, we can include a non-zero vector as a prior mean of genetic effect \nameref{S1_Appendix}. We can exploit information from Mendelian randomization studies of common variants, such as recent findings where rare truncating loss-of-function variants in {\it PCSK9} were found to decrease LDL and triglyceride levels and decrease CAD risk\cite{cohen2005low,pcsk9,do2013common,cohorts2014loss} to identify situations where such a prior is warranted.

@@

\subsection*{UK Biobank Data}
\subsubsection*{GWAS Summary Statistics}
We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described \cite{DeBoever179762}. For asthma, we used the Firth fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in (Hill 2017), but switches to a port of $\texttt{logistf()}$ (\url{https://cran.r-project.org/web/packages/logistf/index.html}) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps. We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array. For variants that were specific to one array, we did not use array as a covariate.

>>>>

 To define the alternative model, we must specify the prior correlation structure, scale, and location (direction) of the effect sizes. Let $N$ be the number of individuals and $K$ the number of phenotype measurements on each individual. Let $M$ be the number of variants in a testing unit ${\bs G}$, where ${\bs G}$ can be, for example, a gene, pathway, or a network. Let $S$ be the number of studies where data is obtained from - this data may be in the form of raw genotypes and phenotypes or summary statistics including linkage-disequilibrium, effect sizes (or odds ratio), and standard error of the effect size. When considering multiple studies ($S > 1$), multiple rare variants ($M > 1$), and multiple phenotypes ($K > 1$), we define the prior correlation structure of the effect sizes as an $SMK\times SMK$ matrix $\mathbf{U}$. In practice, we define $\mathbf{U}$ as a Kronecker product, an operation of matrices of arbitrary size, of three sub-matrices:
\begin{itemize}
\item an $S\times S$ matrix $\mathbf{R_{\textrm{study}}}$ containing the correlations of genetic effects among studies where different values can be used to compare different models of association, such as for identifying heterogeneity of effect sizes between populations~\cite{band2013imputation};
\item an $M\times M$ matrix $\mathbf{R_{\textrm{var}}}$ containing the correlations of genetic effects among genetic variants, which may reflect the assumption that all the PTVs in a gene may have the same biological consequence\cite{macarthur,rivas2013assessing,rivas2015effect} or prior information obtained through integration of additional data sources, such as functional assay data\cite{majithia2014rare,findlay2014saturation}, otherwise zero correlation of genetic effects may be assumed, which is used in dispersion tests like C-alpha\cite{calpha,clarke2013flexible} and SKAT\cite{skat}; and
\item a $K\times K$ matrix $\mathbf{R_{\textrm{phen}}}$ containing the correlations of genetic effects among phenotypes, which may be obtained from common variant data\cite{cotsapas2011pervasive,solovieff2013pleiotropy,gencorr2015}.
\end{itemize}
The variance-covariance matrix of the effect sizes may be obtained from readily available summary statistic data such as in-study LD matrices, effect size estimates (or log odds ratios), and the standard errors of the effect size estimates (\nameref{S1_Appendix}).

@@

Similarly, we can utilize a prior on the location (direction) of effects to specify alternative models where we seek to identify variants with protective effects against disease. Thus far we have assumed that the prior mean, or location, of genetic effects is zero which makes it feasible to analyze a large number of phenotypes without enumerating the prior mean across all phenotypes. To proactively identify genetic variants that have effects that are consistent with a protective profile for a disease, we can include a non-zero vector as a prior mean of genetic effect (\nameref{S1_Appendix}). We can exploit information from Mendelian randomization studies of common variants, such as recent findings where rare truncating loss-of-function variants in {\it PCSK9} were found to decrease LDL and triglyceride levels and decrease CAD risk\cite{cohen2005low,pcsk9,do2013common,cohorts2014loss} to identify situations where such a prior is warranted.

@@

\subsection*{UK Biobank Data}
\subsubsection*{GWAS Summary Statistics}
We performed genome-wide association analysis using PLINK v2.00a(17 July 2017) as previously described \cite{DeBoever179762}. For asthma, we used the Firth fallback in PLINK, a hybrid algorithm which normally uses the logistic regression code described in \cite{hill2017stepwise}, but switches to a port of $\texttt{logistf()}$ (\url{https://cran.r-project.org/web/packages/logistf/index.html}) in two cases: (1) one of the cells in the 2x2 allele count by case/control status contingency table is empty (2) logistic regression was attempted since all the contingency table cells were nonzero, but it failed to converge within the usual number of steps. We used the following covariates in our analysis: age, sex, array type, and the first four principal components, where array type is a binary variable that represents whether an individual was genotyped with UK Biobank Axiom Array or UK BiLEVE Axiom Array. For variants that were specific to one array, we did not use array as a covariate.
